Introduction_outline

### 2019-01-18 Write_up/Writing/Introduction_outline.Rmd
Improved intro

### Introduction

Seawater temperature is a key indicator of environmental change in both terrestrial and marine ecosystems (Pearce et al., 2006), which directly influences the earth’s natural ecological functioning. Climate change is widely believed to result in a rise in the earth’s average surface temperatures over time (Stocker et al., 2013) and this is usually associated with an increase in the intensity and duration of extreme events such as cold-spells, heatwaves, floods and storms. Consequently, these extreme climatic events can often have profound short-term effects on both humans as well as many ecosystems (Easterling et al., 2000).

‘Cold spells’ are usually referred to as a phenomenon where abnormally cold water remains on the surface of the ocean for extended periods of time. However, several other definitions exist based on the statistical properties of the temperature record in association with the location of temperature collections and the time of year .  Studies that have quantified the impacts of extreme temperatures have provided evidence of demonstrably negative effects on  associated surrounding regions as well as the specific regions within which these climatic events occur. While extensive research has been done on other extreme weather events such as floods, wind storms and marine heat waves, minimal research has been done on marine cold spells (MCS). It is known that warm-water anomalies may result in severe coral bleaching and is one of the factors resulting in the demise of coral reefs. Additionally, anomalous warm water events are known to have negative impacts on species diversity and mortality as well as the distribution of organisms. However, recent studies done by Lirman et al. explains that corals living in regions or surrounding regions of unusually prolonged anonymously cold waters may be just as detrimental; this was illustrated in January of 2010 when the reefs of the Florida Reef Track (FRT) were exposed to an unusual cold water event that resulted in large scale coral mortality. This climatic event showed clear association with North Atlantic Oscillation (NAO) index that produced northerly surface wind anomalies and the southward advection of the cold Arctic air. Similarly, research by Shinn et al. (1962) reported high mortality of Acropora cervicornis in shallow reefs as temperatures dropped. Likewise, studies conducted by Hudson et al.  also documented the mortality of Montastraea colonies in near shore patch reefs due to cold fronts between 1969–1970 and 1977–1978, respectively.

MCSs are lethal to all marine organisms and are known to have caused  mass extinctions of several fish and invertebrate species (Woodward. 1987 & Gunter. 1941, 1951 & Holt and Holt. 1983, Gunter. 1951 & Crisp. 1964). Temperature has long been considered one of the primary factors controlling species distribution patterns by limiting their distribution north or southward towards the high latitudes (Firth et al., 2011). Studies done by Firth et al. (2011, 2015) have demonstrated the range shifts of engineer species such as mussels as a result of MCSs. Thus, providing a clearer understanding of the influence of oceanic and atmospheric processes on sea surface temperatures (SST) within the upper mixed layer, however minimal research has been done on the influences of local and broad scale impacts on seawater temperature.

Research by Grunter (1941) and Firth et al. (2011) explains that some MCSs that are believed to have influenced species populations were caused by atmospheric cold spells which had ultimately affected the local intertidal and coastal biota. MCSs have been proven to have a great effect on ecosystem change, thus understanding these climatic drivers may provide useful tools for conservation and management efforts. This study will aim to develop a better understanding of the impacts of weak and strong upwelling events and how they may be influencing the intensity and duration of MCSs. This study applied the MCSs definitions to datasets at a local-scale and an offshore broad scale.

Eastern Boundary Upwelling Systems (EBUS) are among the most productive regions of the world. These upwelling systems include the; California, Humboldt, Canary and Benguela currents, and are important for biological productivities by supporting commercial fisheries (Pauly and Christensen, 1995) thus forming an essential component to these region’s economies, as well as providing recreational services to the human population along their coasts. One of the five continental margin upwelling systems is the Benguela Current; an eastern boundary current of the south Atlantic subtropical gyre which flows towards the equator approximately 300 km offshore of the coast of southwest Africa before splitting into the Benguela Coastal Current (BCC) and Benguela Oceanic Current (BOC). Shoreward of the main stream of the Benguela current there is a slow flow along the continental margin into the Angola Basin. In this slow flow the prevailing south and south eastern winds drive coastal upwelling of cold, nutrient rich South Atlantic Central Water (SACW).

The SACW originates by sinking at the Subtropical Front, known as the Subtropical Convergence (Schell, 1970 & Nelson, 1989). Beneath the SACW lies cold, nutrient-rich Antarctic Intermediate Water (AAIW) which originates by sinking at the Polar Front. Throughout each of the seasons, upwelling of nutrient rich waters covers most of the continental shelf and much of the continental slope (Shannon, 1985), thus creating what we refer to as the Benguela Current Upwelling System. Research studies have shown that these cold waters move north over the shelf in the Benguela Coastal Current (Hay and Brock, 1992). Offshore they are separated from the warmer waters of the subtropical Atlantic by a well-developed thermal front (Shannon, 1985). This quasistationary region is often linked with the shelf edge, although it is distorted by eddies, plumes and jets which extend west into the SE Atlantic (Lutmharms and Stockton, 1987; Waldron and Probyn, 1992). These plumes largely expand the region of high productivity as a result of upwelling by rapidly advecting cold, nutrient-rich water from the coastal upwelling sites into the open ocean (Lutjeiarms et al, 1991). Northward, the Benguela Current Upwelling system is linked by the Angola Benguela Front which converge between the Benguela current and the Benguela coastal current as well as the warm saline waters of the Angola Current (Peterson and Stramma, 1991). Southward, the coastal upwelling system is bounded by the Agulhas Bank. South of the African coast the warm Agulhas Current flows in from the Indian Ocean and returns eastwards while generating warm eddies that spin off in a north-westerly direction in the Benguela current (Duncombe, 1991). Research studies done by Shannon et al. (1985) explains that coastal upwelling varies on a seasonal basis and that during winter and spring, water of approximately 16°C extends along the coast. Upwelling is coldest and most persistent off the Namibian coast in the Lüderitz region at 25°S and weakest at 17°S (Lutjeharms and Meeuwis, 1987). Lutjeharms and Meeuwis, (1987) also found a strong relationship between the intensity of upwelling events and the strength and direction of prevailing winds.  It is also hypothesised that upwelling is more intense where deep water is closer to the coast.

EBUS mesoscale and sub-mesoscale oceanic movements are known to influence marine ecosystems in that it enhances primary production (McGillicuddy et al., 1998; Oschlies and Garçon, 1998) and is key in managing the exchange and of transporting of organic resources and marine organisms from the coastal zone to the open ocean. Large-scale atmospheric pressure systems that favour along-shore, equatorward winds, result in the high levels of productivity in the EBUS, which in combination with the Coriolis effect, affects surface water offshore. This results in a process whereby cold, nutrient rich, with low pH and low oxygen concentration, originates from deep within the ocean and is pushed to the surface where they, in the form of phytoplankton blooms, fuel primary production. As a result of the intensity and duration of the upwelling event; regions affected by these cold water events are exposed to distinct ecological conditions and tend to differ from its surrounding regions (Bosman et al., 1987; Barshis et al., 2011). Significant research evidence suggests that SST have increased, however some regions still show a lower or non-significant increase, and in some cases cooling trends are evident. Inshore regions and embayment’s that are protected from wind action may be subjected to intense episodes of primary production (Graham and Largier, 1997). Vander Wouder et al. (2006) found that these sites represent the highest levels of chlorophyll concentration.

Atmospheric and oceanographic processes results in upwelling, while the amplitude and timing of upwelling-favourable winds are sensitive to climate variability (Montecinos et al., 2003; Macias et al., 2012), and are highly likely to be affected by global warming (Barros et al., 2014; Bakun et al., 2015). It is hypothesised that global warming will influence atmospheric pressure gradients, ultimately affecting the amplitude and timing of coastal winds that cause upwelling (Bakun, 1990; Bakun et al., 2015) and more recent observations (Rykaczewski et al., 2015; Wanget al., 2015) by means of global model projections strongly suggests intensification of coastal winds. Understanding the influence of changing winds on upwelling events is highly important. For example, weak winds may result in weak upwelling, which in turn may limit nutrient concentration; resulting in potentially negative impacts on primary production (Chhak and Di Lorenzo, 2007). In contrast, stronger upwelling winds, could theoretically result in stronger upwelling events, thereby increasing the nutrient concentration which may increase offshore transport (Cury and Roy, 1989). Increased wind action may also result in increased turbulence (Cury and Roy, 1989); and may even affect the processes associated with ocean acidification, creating widespread changes within ecosystems and ultimately impacting regional productivity (Gruber, 2011; Doney et al. 2012).

**** A better understanding of the impact of global warming on wind variability is important in that it allows for better management of fisheries and other marine resources in the EBUS which could be improved with a better understanding of how nutrient abundance and primary production are associated with upwelling events. This study aims to emphasize the drivers of upwelling-favourable winds, as well as the factors influencing nutrient enrichment, and potential changes in biogeochemistry*****

The Benguela current, forming part of the EBUS, experiences cross-shore atmospheric pressure gradients, which results in alongshore, equatorial winds. These winds are known to drive coastal upwelling events. With this in mind, research by Bakun (1990) suggests that an increase in greenhouse gas emissions would exacerbate warm-season upwelling. The understanding behind this mechanism was based on the idea that air above land would warm rapidly as compared to air above the ocean, and that the deepening of the continental thermal low pressure system (CTLPS) would increase the cross-shore pressure gradient which drives upwelling-favourable winds. Several research studies examining the trend between upwelling events and wind patterns, have however indicated contrasting results. With more than 20 years of observational and modelled data; Sydeman et al. (2014a) discovered the most coherent pattern of intensifying upwelling-favourable winds in the Humboldt and California upwelling systems. It has proven difficult to attribute coastal wind intensification in EBUS to global warming, despite the fact that the meta-analysis done by Sydeman et al. (2014a) generally supports Bakun’s (1990) proposition.

This study cannot discount the role of multi-decadal climate variability in the observed trends (Chhak and Di Lorenzo, 2007; Narayan et al., 2010; Pérez et al., 2010; Macias et al., 2012; Santos et al., 2012; Cropper et al., 2014; Jacox et al., 2014), nor is it a test of the Bakun (1990) hypothesis mechanism. Given the broad range of contradicting studies, the only conclusion that can be made is that upwelling-favourable winds tend to be the most intense during the warmer months of the year (Huyer, 1983; Nelson and Hutchings, 1983). 
The timing and the intensity of upwelling favourable winds are highly variable within the EBUS, therefore creating much uncertainty (Chavez and Messié, 2009). Understanding the factors contributing to the intensity and duration of upwelling events is important because variations in the timing of upwelling strongly influences ecosystem productivity. Since the late 1940s studies have shown an increase in upwelling favourable winds associated with the California current, however, less variability has been observed in the strength of upwelling winds through the year in the central Benguela and Canary Currents (Shannon and Nelson, 1996; Varela et al., 2015). It is important to note that the lack of research conducted may obscure the validity of any conclusion drawn here. This study further researched the changes in the timing and variability of upwelling within the Benguela Current and how these phenomena influence associated ecosystems.

As a result of climate change, global ocean temperatures have increased at an accelerating rate (Stocker et al., 2013). Sufficient research suggests that coastal sea surface temperatures (SST) in EBUS such as the Benguela current are cooler than offshore regions as a result of the near shore upwelling waters. There is a large but contradicting amount of evidence on the SST trends in EBUS (Mendelssohn and Schwing, 2002; Arístegui et al., 2009; Belkin, 2009; Demarcq, 2009; Lebassi et al., 2009; García-Reyes and Largier, 2010; Rouault et al., 2010; Gutiérrez et al., 2011; Seo et al., 2012; Barton et al., 2013; Barros et al., 2014; Sydeman et al., 2014b; Salvanes et al., 2015). The contradictions amongst studies exist because of varying resolutions of SST data used, thus resulting in the capturing of different aspects of the upwelling processes, showing different trends that exist based on the resolutions of the different SST datasets as well as other intrinsic factors. Generally, because of the resolution of SST datasets it is often difficult to separate near shore upwelling temperatures from regional, broad scale and eastern boundary current temperatures. Lima and Wethey’s (2012) found that high resolution data represented clear cooling trends predominantly during summer and at high latitudes in most of the EBUSs. Additionally several research studies examined the difference between coastal and offshore temperature trends, revealing decreasing or increasing trends in near shore SST, compared with increasing trends offshore in the Benguela compared with increasing trends offshore in the Benguela (Rouault et al.,2010; Santos et al., 2012; Salvanes et al., 2015). In this study however, will utilise a range of spatio-temporal resolutions of SST data and in situ collected seawater temperature measurements for the Benguela upwelling system. The aim of this is to demonstrate the variable upwelling events based on the different resolutions at which temperature is being collected.

Generally, regional ocean climate oscillations like the Benguela Niño for the Benguela Current System impose considerable variability on temperatures and complicate attempts to assess long-term trends (Demarcq,2009). It is predicted that these climate oscillations may increase the variance with climate change (Timmermann et al., 1999; Kuzmina et al., 2005; Sydeman et al., 2013; Cai et al., 2015) which will further result in unusual trends.

Research studies provide significant evidence on the increase in warming of the ocean surface. This warming slowly propagates down through the water column and in turn increases the thermal stratification (Gruber, 2011). Strong stratification limits the depth at which water is up welled and hence controls the amount of nutrients brought up (Chhak and Di Lorenzo, 2007; Jacox and Edwards, 2011; Jacox et al., 2015). The interplay between upwelling winds and stratification determine the amount of nutrients present within the system and thus is strongly linked to biological productivity. Globally; as a result of climate change, stratification of the water-column between 0 and 200 m has increased by an estimated 4% during the period 1971 to 2010 (Stocker et al., 2013). Variability in the biogeochemical properties of up welled waters influences species distribution, ecosystem structure and productivity within the Benguela upwelling system. The deep waters being pushed to the surface through the upwelling process is high in nutrients and dissolved CO2, a consequence of the long period of isolation at a depth from the euphotic zone (Feely et al., 2008). Increased ocean stratification as a result of surface warming may reduce the ventilation of deep-water masses. As the deep water is being up welled to the surface, decreases in oxygen and pH levels are predicted along with an increase in nutrient levels (Rykaczewski and Dunne, 2010).

Biotic systems within the Benguela upwelling system may be negatively influenced by the changes in oxygen levels given their adaption to natural variable conditions (Capone and Hutchins, 2013).  This study will seek to examine the impact of both strong and weak upwelling winds on the intensity and duration of the upwelling event; additionally it will also aim to examine how chlorophyll concentrations vary with regard to contrasting upwelling conditions as well as the impact this has on species distribution and habitats.

##### Ideas #####

-chnaging the structure it to describe MCS, explain the dangers of these phenomena on marine life (, the implications of this (eg distributions being affected), and the causes of MCS.

-explain what info is unknown about MCS and why this is a problem in areas where they can occur

- link this to the EBUS and Benguela stuff

- clearly mention the aims, objectives and approach to the study.

>>>>

Chapter 1: General Introduction

## Eastern boundary upwelling systems:

Globally, there are four major eastern boundary upwelling systems (EBUS). These are the Benguela Current off south-western Africa, the Canary Current off north-western Africa including its northern extension off the Iberian Peninsula of south-western Europe, the Peru-Humboldt Current off western South America, and the California Current off the western continental USA and north-western Mexico (Bakun et al, 2015; Pauly and Christensen, 1995). Each of these systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans. EBUS cover only 1% of the world’s ocean surface, but despite these are among the most productive regions of the world. This is because these systems are subject to coastal upwelling, which is defined as the process whereby cold, nutrient rich, high CO2 concentrated, low pH, and low oxygen waters are pushed to the surface as a result of alongshore winds interacting with the earth’s rotation. As a result of upwelling, diverse marine fauna and flora are able to thrive in these areas because of the abundance of nutrients and favourable temperatures.

Upwelling is primarily caused by alongshore, equator ward winds. These winds are caused by cross-shore atmospheric pressure gradients, and these gradients occur predominantly during heating periods. Bakun then proposed that if there is an increase in greenhouse gases, it will cause more heating during the day and more cooling during the night which cause the temperature gradient to increase and would result in a strong pressure gradient, and as we know the pressure gradient will affect the wind and therefore will affect upwelling. Several people have tried to model this using different datasets of SST and wind pattern data at various resolution and initially found that increases in gases, such as CO2 causes a reduction in trade winds, and therefore a reduction in upwelling. However, these models were based on coarse data and when higher resolutions are used, such as for the Benguela EBUS, the models have shown that upwelling will intensify in accordance to the upwelling intensification hypotheses (eg, Hsieh and Boer 1992, Mote and Mantua 2002, Lima and Wethey’s 2012).

### Ecology and primary production within upwelling zones:

The duration and intensity of coastal upwelling are known to have a critical role in the phenology of key marine ecosystem processes. Similarly, changes in the characteristics of upwelling are shown to have cause large disturbances to ecosystems at several trophic levels (Wang et al., 2015). Cold, nutrient rich, low pH, and low oxygen upwelled waters fuels primary production as a consequence of the development of phytoplankton blooms (Ryther 1969, Cushing 1971, Payne 1987, McGillicuddy et al., 1998; Oschlies and Garçon, 1998). In these upwelling systems, the high productivity supports vast and diverse marine populations. Mesoscale and sub-mesoscale oceanic movements are key in managing the exchange and transportation of organic resources and marine organisms from the coastal zone to the open ocean. Most of the world’s major fishing industries are dependent on upwelling, because this process brings nutrient rich waters into the euphotic zone, where it forms the foundation of a localized food web that ecologically differs from surrounding, non-upwelled regions (Bosman et al., 1987; Barshis et al., 2011). Whilst nutrients play a primary role in the survival of these systems, the cool temperatures of these waters are critical as well. There is an increasingly large body of research that has provided evidence showing that sea surface temperatures (SST) have and are continuing to increase. However, some regions still show a lower or non-significant increase, and in some cases cooling trends are evident (REFERENCES). Upwelling occurring along the Cape south coast is proven to have less of an economic importance in contrast to the west coast. Inshore regions and embayments that are protected from wind action may be subjected to intense episodes of primary production (Graham and Largier, 1997). Vander Wouder et al. (2006) found that these sites represent the highest levels of chlorophyll concentration. Biotic systems within the Benguela upwelling system may be negatively influenced by the changes in oxygen levels given their adaption to natural variable conditions (Capone and Hutchins, 2013).

For several decades, regime shifts have become an important factor of marine research. Regime shifts are often defined as a shift from one stable ecosystem state to another. This process affects several species within a short period of time (deYoung et al., 2004; Cury and Shannon, 2004; Jarre et al., 2006, Blamey 2012). Regime shifts are thought to be driven by abiotic processes such as climate change (Polovina, 2005, Blamey 2012). The southern Benguela upwelling system supports a wide variety of commercially-exploited fish species and exhibits substantial variability in both oceanographic and biological components. Over the past 3 three decades, a number of species occurring within this region have undergone shifts in their distributions and Roy et al. (2007) linked these shifts to cooling waters along the inner shelf of the Agulhas bank on the south coast, east of Cape Agulhas (Blamey, 2012). This coast is associated with high levels of harmful toxic algal blooms, which negatively impact commercial and recreational interests along the coastal region. These harmful blooms are often attributed to flagellated species with the red tide representing dinoflagellate blooms (Pitcher et al. 1993A, Pitcher 1998). Formations of red tides are often related to the prevailing winds of the southern Benguela which govern most hydrodynamic processes on the continental shelf. Red tide events are greatest during the latter part of the upwelling season during transitions in the synoptic weather patterns which result in diminished upwelling activity and increased thermal stratification (Pitcher et al. 1993a).

### Atmosphere:

Throughout region surrounding the Benguela EBUS, the atmosphere is dominated by anticyclonic high pressure cells with quasi-stationary positions. The South Atlantic Ocean High is situated along the west, drawing cool, dry air onto the west of the subcontinent (Van Heerden and Hurry, 1998). Solar heating during summer may result in the development of low pressure cells, known as heat lows. These cells are absent during winter time (Tyson and Preston-Whyte, 2000). When the anticyclones shift north, during winter months, the cold westerlies have a substantial impact on the weather of the southern tip of the South African subcontinent (Van Heerden and Hurry, 1998). The atmospheric temperatures along the coastline are largely influenced by the Benguela Current (Van Heerden and Hurry, 1998). However, upwelling systems may be forced by various atmospheric conditions (Lutjeharms et al., 2003; Roberts, 2005; Hutchings et al., 2009).

### Sea surface temperature and data collection:

Water temperature affects the survival, abundance, distribution and growth of marine species (REFERENCE). As such, a clear understanding of the relationship between marine ecosystems and their thermal tolerance is necessary for conservation and management planning. Future predictions in biogeographical patterns and ecosystem functioning may also be estimated based on various climatic behaviours. Regardless of the objectives, whether it be studies based on climate change, extreme events or ecosystem functioning, it is important that the research is conducted through accurate, reliable datasets. Unfortunately, not all data are created equally and under the same conditions and as such not all datasets are usable for a particular task.

Remotely-sensed satellite SST products are produced and maintained by a number of organisations worldwide. In order to obtain these SST data, advanced technology is launched into space, in order to obtain infra-red light being emitted by the ocean. This irradiance is then taken as a proxy for SST.  Due to the distance at which these SST are being collected, certain coarseness within the datasets are unavoidable. Satellite SST products undergo a range of quality control processes that are well researched and tested (Reynolds and Smith, 1994; Brown et al., 1999; Martin et al., 2012), making their use more reliable. However, Smit et al. (2013) have shown that SST data are not particularly reliable for coastal regions and should be used cautiously as they can be prone to both warm and cold bias. Nevertheless, satellite data products are extremely useful when investigating the ocean at broad scale and can provide important SST data.  <will add more here on the different types of resolution used in this thesis>

### Global warming and climate change:

Climate change as a result of anthropogenic activities is now established as a global concern for both scientific and political importance. As the human population grows, the rate of global warming increases, resulting in profound effects on marine and other ecosystems. Climbing temperatures create a host of additional changes to marine systems, such as rising sea levels, increased ocean stratification, decreased sea-ice extent, and altered patterns of ocean circulation and precipitation (Doney et al., 2011). Furthermore, direct effects on ocean temperature changes and chemistry may alter the functioning and behaviour or marine organisms. In turn, these changes lead to a change in species integration and trophic pathways.

Significant research studies accept the climate change and variability have great impacts on marine ecosystems (Stenseth et al. 2002, Harley et al. 2006, Hoegh-Guldberg & Bruno 2010). The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of greenhouse gas concentrations (Solomon et al. 2007). A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased largely. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean thus presenting regions warming much faster or slower than then the global average (Levitus et al., 2000; Palttridge and Woodruff, 1981; Santos 2012). In particular, the Atlantic Ocean contributes most to the increase of the heat content (Nerem et al., 1999; Levitus et al., 2000; Strong et al., 2000).

As a result of climate change, research studies provide significant evidence on the increase in warming of the ocean surface. This warming slowly propagates down through the water column and in turn increases the thermal stratification (Gruber, 2011). Strong stratification limits the depth at which water is upwelled and hence controls the amount of nutrients brought up (Chhak and Di Lorenzo, 2007; Jacox and Edwards, 2011; Jacox et al., 2015). The interplay between upwelling winds and stratification determine the amount of nutrients present within the system and thus is strongly linked to biological productivity. Globally, as a result of climate change, stratification of the water-column between 0 and 200 m has increased by an estimated 4% between 1971 to 2010 (Stocker et al., 2013). Increased ocean stratification as a result of surface warming may reduce the ventilation of deep-water masses. As the deep water is being up welled to the surface, decreases in oxygen and pH levels are predicted along with an increase in nutrient levels (Rykaczewski and Dunne, 2010).  Generally, regional ocean climate oscillations like the Benguela Niño for the Benguela Current System impose considerable variability on temperatures and complicate attempts to assess long-term trends (Demarcq, 2009). It is predicted that these climate oscillations may increase the variance with climate change (Timmermann et al., 1999; Kuzmina et al., 2005; Sydeman et al., 2013; Cai et al., 2015) which will further result in unusual trends.

As previously mentioned, upwelling winds in EBUSs are generated by pressure differences between continental thermal lows and oceanic highs. As warming intensifies, pressure gradients between land and sea increases, resulting in a greater intensity of upwelling winds (Bakun et al. 2010). In rare cases, projected upwelling increases may overcome the countervailing effects of upper-ocean warming and stratification to cause regional cooling (Auad et al.2006). Research studies reveal declines in ocean oxygen levels as a result of ocean warming and reduced ventilation from stratification and circulation changes (Keeling et al. 2010). Similarly, low oxygen levels are also possible because of shifts in the composition of sinking organic material (Hofmann & Schellnhuber 2009, Doney et al., 2011).

## Problem statement:
With global warming continuing at a rapid pace the amount of greenhouse gases within the atmosphere are going to increase in the future. This increase will undoubtedly have an impact on the world’s oceans and the marine life within, both directly and indirectly. The oceanic process of coastal upwelling is one aspect that is likely to be affected as changes in global temperature will affect temperature and pressure gradients, and in turn affect alongshore winds which are responsible for driving this process within EBUS regions. Because upwelling is important for productivity within coastal systems, a decrease in this process will negatively impact on marine fauna and flora, and on human activities such as fishing. To assess changes in upwelling events, here I aim to use SST, wind, chlorophyll data etc. to monitor durations and intensities of upwelling within one of the EBUS.....

## Study Area:

This study focusses on the Benguela EBUS. The Benguela EBUS is unique in that it is bounded at both the equator ward and pole ward ends by warm water regimes. These currents are forced locally by the wind stress field off South-West Africa (Nelson and Hutchings 1983; Fennel 1997). This upwelling system stretches along the south-western coast of Africa from southern Angola to Cape Agulhas (Cole, 1999). Water is coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S (Lutjeharms and Meeuwis, 1987). The main upwelling period in the Benguela north of 25°S extends mainly from March to November, peaking during August (Shannon, 1985), but upwelling persists throughout the year. Upwelling season in the northern Benguela extends between September and March, with very little upwelling taking place during winter (June-August). In the central Benguela and also at Cape Frio upwelling is continuous throughout the year (Shannon, 1985). The cold Benguela waters are bound in the south by the warm Agulus retroflexion region and in the north by the southward flowing Angola current. The flow of the Benguela is often considered to be topographically guided (Nelson and Hutchings, 1938; Barange et al. 1992). The Benguela system is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region (Shannon 1985; Cole 1999).

Upwelling in the northern Benguela is in phase with the seasonal insulation cycle, whereas in the southern Benguela it is not. The result is that there is a pronounced seasonal cycle in SST in the north while in the south there is no clear SST cycle.  The seasonal warming occurring during the late summer season off northern and central Namibia is a regular occurrence and is associated with warmer, higher salinity water of Angolan origin into the region (O'Toole, 1980; Shannon, 1985). There have been a number of warm and cool periods documented in the northern and southern Benguela (e.g. Stander and De Decker, 1969; Walker et al.1984; Shannon et al., 1984; Boyd and Thomas, 1984; Shelton et al. 1985; McLain et al.: 1985). Walter (1937) found a major climatic event in the northern Benguela region of anonymously warm water. This unusual warm water may have resulted due to reduced upwelling or an increase southward advection of warm water. Another warming event occurred during the summer and autumn months of 1984 (Shannon et al, 1986). The presence of this anomalous warm water during 1984 followed two cooler than normal years in the northern Benguela. Shannon et al. (1986) illustrated the warm (21-23°C) water near Walvis Bay as a large warm water patch offshore west of Luderitz. The absence of upwelling between Walvis Bay and north of Luderitz is evident however, the upwelling centre in the Benguela system (Shannon, 1985), and upwelling was minimal. Interestingly, in the southern Benguela region, there was no evidence of any unusual patterns in temperature distributions. The seasonal temperature pattern within the northern Benguela is somewhat similar to that in the eastern equatorial Atlantic and the regular southward intrusion of warm water into the northern Benguela during the summer season. During these warming periods the equator ward wind stress over the northern Benguela region was stronger than normal; yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region.

Northward, the Benguela Current Upwelling system is linked by the Angola Benguela Front which converge between the Benguela current and the Benguela coastal current as well as the warm saline waters of the Angola Current (Peterson and Stramma, 1991). Research done by Yamagata and Iizuka (1995) discovered slight indication that coastal Kelvin waves, originating from equatorial Kelvin waves, carry the down welling process southward.  Southward, the coastal upwelling system is bounded by the Agulhas Bank. South of the African coast the warm Agulhas Current, western boundary current, flows in from the Indian Ocean and returns eastwards while generating warm eddies that spin off in a north-westerly direction in the Benguela current (Duncombe, 1991). The region surrounding the Benguela experiences a persistent along shore wind. This wind is associated with the St. Helena high pressure system. Lutjeharms and Meeuwis, (1987) also found a strong relationship between the intensity of upwelling events and the strength and direction of prevailing winds. These favourable upwelling equatorial winds represent a maximum at 25°S and decrease towards the northern and southern boundaries of the Benguela upwelling system. Continental shelf bathymetry and upwelling favourable winds provide a large-scale upwelling mechanism in the southern Benguela, whereas local topography and meteorology create an alternating pattern of active and passive upwelling circulations along the coast. Winds occurring within the southern regions represent a seasonal variation, reaching a maximum during spring and summer months (Boyd 1987; Shannon and Nelson 1996; Fennel 1997). Winds occurring within this region tend to show increasing patterns away from the coast whereas, winds occurring in the northern region exhibits relatively less seasonal variation (Shannon 1985. Pitcher et al. 1992). Upwelling occurring along the South Coast of the Cape Province within South Africa, specifically along the coastline is generated by local wind.  The wind increases somewhat away from the coast. This facilitated the idea of wind stress curl, as discussed in Bakun and Nelson (1991) and Shannon and Nelson (1996).

## Extreme events

### Cold spells
- Definition
- Impact
-Major cold spells

## Aims and approach:
In this thesis I aim to…
